Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-21 paper/paper.tex
Intro redone.... again

\section{Introduction}
\label{sec:intro}
Structured recursion schemes are powerful abstractions that capture common
patterns of recursion. The main benefit of structuring computation using
recursion schemes, is that they enjoy well-established \emph{algebraic
properties} that can serve as a foundation for reasoning about program
equivalences, transformations, and optimisations (e.g.\ \emph{fusion} laws, or
semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
%
These algebraic properties are widely used in the context of \emph{program
calculation}, where programmers describe their code using simple, inefficient
specifications in an \emph{algebra of programming}~\cite{BirddeMoor96:Algebra},
and then use the algebraic laws of this algebra of programming to
\emph{calculate} an efficient version of the same algorithm.
%
Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
implementation, with
\ocaml{map (fun x -> x * 2)}.
A lot of program calculations have this pattern: start from a simple
specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
an optimised version. We will revisit a similar example in
Section~\ref{sec:sorting}.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is a
\emph{functor} $\Set \to \Set$ mapping objects and arrows in $\Set$ to $\Set$
itself (such functors from a category to itself are called \emph{endofunctors}).
Its functorial action $\List{f} : \List{A} \to \List{B}$ is given by
$\List{f}(\nil(*)) = \nil(*)$ and $\List{f}(\cons(a,xs)) = \cons(f(a),
\List{f}(xs))$. Notice that the definition $\List{f}$ is well-defined as it
recursively calls on a smaller argument.  Similarly, the set of streams $\Str{A}
\iso A \times \Str{A}$ is the greatest set generated by the constructor $\cons :
A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and $\tail :
\Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.

@@

\section{Extraction}
\label{sec:extraction}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic functional code. Although our examples use OCaml as target
language, extraction works equally well to other target languages. Our examples
are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynamic
programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
deforestation optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
We compare our work with other
encodings of program calculation techniques and recursion schemes in Coq, and
discuss other approaches to fusion and to termination of nonstructural
recursion.
% and recursion schemes in Coq
% (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
% encoding of recursion schemes to other forms of termination checking for
% nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
% approaches to fusion in functional programs (Section~\ref{sec:fusion}).

% \textbf{Program Calculation}.
% \label{sec:calccoq}
Within the domain of program optimisation, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

@@

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

>>>>

\section{Introduction}
\label{sec:intro}
Over the years, extensive research has been conducted on program calculation
techniques, a well-established approach for deriving efficient programs from
simpler specifications through systematic program
transformation~\cite{gibbonsSquiggol, BirddeMoor96:Algebra}. A key aspect of
this approach is the use of structured recursion schemes, which serve as
powerful abstractions that capture common patterns of recursion. By structuring
computation in this way, programs benefit from well-established \emph{algebraic
properties} that provide a solid foundation for reasoning about program
equivalences, transformations, and optimisations -- such as \emph{fusion} laws or
semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}.
In the context of \emph{program calculation}, these algebraic properties allow
programmers to describe code using simple, inefficient specifications within an
\emph{algebra of programming}~\cite{BirddeMoor96:Algebra}. They can then apply
algebraic laws to systematically \emph{calculate} more efficient versions of the
same algorithms.

Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
implementation, with
\ocaml{map (fun x -> x * 2)}.
A lot of program calculations have this pattern: start from a simple
specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
an optimised version. We will revisit a similar example in
Section~\ref{sec:sorting}.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is a
\emph{functor} $\Set \to \Set$, in particular, an \emph{endofunctor}, mapping objects and arrows in $\Set$ to $\Set$
itself.
Its functorial action $\List{f} : \List{A} \to \List{B}$ is given by
$\List{f}(\nil(*)) = \nil(*)$ and $\List{f}(\cons(a,xs)) = \cons(f(a),
\List{f}(xs))$. Notice that the definition $\List{f}$ is well-defined as it
recursively calls on a smaller argument.  Similarly, the set of streams $\Str{A}
\iso A \times \Str{A}$ is the greatest set generated by the constructor $\cons :
A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and $\tail :
\Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.

@@

\section{Extraction}
\label{sec:extraction}
In this section we show how programs and their optimisations can be encoded in
our framework and how can they be extracted to idiomatic functional code.
Although our examples use OCaml as target language, extraction works equally
well to other target languages (e.g. Haskell or Lisp). In particular, we show how to formalise
 quicksort (Section~\ref{sec:sorting}), we give an example of dynamic
programming algorithm, that is knapsack (Section~\ref{sec:knap}), and show examples of the shortcut
deforestation optimisation (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
% and recursion schemes in Coq
% (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
% encoding of recursion schemes to other forms of termination checking for
% nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
% approaches to fusion in functional programs (Section~\ref{sec:fusion}).

% \textbf{Program Calculation}.
% \label{sec:calccoq}

@@

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
Fusion in functional programming is a more general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.
